---
title: "Diamonds Are Forever (Week 6 Homework)"
author: "Nick Mattson"
format:
  html:
    theme: darkly
    code-fold: true
    code-summary: "Show code"
editor: visual
---

```{r}
#| label: setup
#| include: false

# Load Libraries
library(tidyverse)
library(here)
library(ggplot2)
library(kableExtra)

# Let's use the diamonds dataset
diamondsdata <- ggplot2::diamonds

diamonds_filtered <- diamondsdata |>
  filter(carat <= 3) # apply filter to reduce dataset for analysis

diamonds_mean_price <- diamonds_filtered |>
  group_by(cut, carat) |>
  summarise(mean_price = mean(price, na.rm = TRUE)) #summary table
```

The scatterplot below shows the average price in dollars of the `r n_distinct(diamonds_mean_price$cut)` standard cuts of diamond sold on the market, adjusted to carat size. In the original dataset, `r nrow(diamondsdata)` individual gemstones were observed to create this dataset. To remove outliers and make the data more interpretable, the analysis is limited to stones less than 3 carats. This reduced the dataset to `r nrow(diamonds_filtered)` observations.

```{r}
#| label: fig-diamond_plot
#| out-width: "75%"
#| fig-cap: Average Diamond Price by Cut

#scatterplot of gemstones
ggplot(diamonds_mean_price, aes(x = carat, y = mean_price, color = cut)) +
  geom_point(alpha = 0.5) +
  theme_grey() +
  labs(x = "Carat Size",
       y = "Average Price (Dollars)",
       color = "Carat Size",
       caption = "Data from ggplot2 Diamonds package") + 
  scale_color_discrete() +
  theme(axis.title = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 8, face = "italic"))
```

As you can see in @fig-diamond_plot, there is disagreement on how to properly price stones of larger carat size. For stones under one carat, there is a much stronger correlation between carat size and average price. As carat size increases beyond one carat, there is much less consensus on how to accurately price stones.

```{r}
#| label: tbl-diamond_variance_one_carat
#| out-width: "75%"
#| tbl-cap: Average Price and Variance for Diamonds 1 Carat or Less

#table showing values for smaller gemstones
diamonds_filtered |>
  filter(carat <= 1) |>
  group_by(cut) |>
  summarise(mean_price = mean(price, na.rm = TRUE),
            price_sd = sd(price, na.rm = TRUE)) |>
  kbl(col.names = c("Cut", "Avg. Price (\\$)", "Standard Deviations"),
      align = "lrr",                                
      digits = 2,
      escape = FALSE) |>
  kable_styling(full_width = FALSE) |>
  row_spec(color="white",background="transparent",0, bold = TRUE)
```

Note that the number of standard deviations in @tbl-diamond_variance_one_carat are lower than those in Table 2 below, which shows average price and standard deviations for stones greater than 1 carat.

```{r}
#| label: tbl-diamond_variance_multiple_carat
#| out-width: "75%"
#| tbl-cap: Average Price and Variance for Diamonds Larger than 1 Carat

#table showing values for larger gemstones
diamonds_filtered |>
  filter(carat >= 1) |>
  group_by(cut) |>
  summarise(mean_price = mean(price, na.rm = TRUE),
            price_sd = sd(price, na.rm = TRUE)) |>
  kbl(col.names = c("Cut", "Avg. Price (Dollars)", "Standard Deviations"),
      align = "lrr",                                
      digits = 2,
      escape = FALSE) |>
  kable_styling(full_width = FALSE) |>
  row_spec(color="white",background="transparent",0, bold = TRUE)
```

As you can see by comparing @tbl-diamond_variance_one_carat with @tbl-diamond_variance_multiple_carat, stones above one carat have a much larger distribution, indicating disagreements in the market on how to price such stones.

```{r}
#| label: fig-diamond_variance_plot
#| out-width: "75%"
#| fig-cap: Price Distribution by Cut and Carat Class
#| fig-height: 10
#| fig-width: 8

# Create new dataframe splitting diamonds above and below 1 carat
diamonds_with_classes <- diamonds_filtered |>
  mutate(carat_class = if_else(carat <= 1, "1 Carat or Less", "1 Carat or More"))

# Ggplot histogram showing distribution of gemstones
ggplot(diamonds_with_classes, aes(x = price, fill = cut)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.8) +
  facet_grid(cut ~ carat_class, scales = "free_y") + 
  scale_fill_viridis_d(option = "viridis") +
  labs(subtitle = "Comparing Diamonds ≤ 1 carat vs. ≥ 1 carat",
    x = "Price (Dollars)",
    y = "Number of Diamonds",
    fill = "Cut"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10), 
    legend.position = "none"
  )
```

In order to ensure a good price on larger stones (above 3 carats), hiring a gemstone expert is recommended.
